---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Example 10.14: Spatial modeling of malaria in Gambia {-}

```{r Ex 10.14 clean, include = FALSE}
rm(list=ls())
```
Note: There might be a warning when loading the $\texttt{raster}$ package, if necessary, uninstall and re-install the package. 

For this example we are using the same $\texttt{gambia}$ data set from the $\texttt{geoR}$ package but an $\texttt{id_area}$ column was added using QGIS to differentiate the different areas as in the original paper. Therefore, we need to load the [gambia_area.csv](https://github.com/spacetime-environ/stepi2/blob/main/data/malaria_gambia/gambia_area.csv) file.

```{r Ex 10.14 load libraries and data, message = FALSE, warning=FALSE, error=FALSE}

library(dplyr) # to manipulate the data
library(geoR) # to get the dataset
library(ggmap) # to plot the map
library(nimble) # for modeling
#library(raster) # to get the environmental data
# library(rgdal) # for adding and transforming coordinates
library(sf) # manipulate spatial data
library(sp) # for manipulating spatial data
library(stringr) # to analyze posterior
library(viridis) # for a more cheerful color palette

# Note, we are using the same Gambia dataset as in the geoR package but an id_area
# attribute has been added using QGIS to a
gambia <- read.csv("MalariaGambia/gambia_area.csv") # gambia dataset from geoR package

```

Since the data is given at the individual level, we want to aggregate the malaria tests by village. 
If we explore the data frame we see that there are 2035 individuals at 65 villages. 

```{r Ex 10.14 explore data}
head(gambia)
dim(gambia)
dim(unique(gambia[, c("x", "y")]))
```

We create a new data frame aggregated by village containing the coordinates, the number of malaria tests, and the prevalence. 

```{r Ex 10.14 aggregate data}

malaria_village <- group_by(gambia, x, y) |>
  summarize(total = n(),
            positive = sum(pos),
            prev = positive / total) |> as.data.frame()

head(malaria_village)
```

```{r Ex 10.14 plot prevalence, message = FALSE, warning=FALSE}

# create a new variable in "sp" format and define coordinates
malaria_utm <- malaria_village
coordinates(malaria_utm) <- ~ x + y
proj4string(malaria_utm) <- CRS("+proj=utm +zone=28")
# convert to long lat 
malaria_geo <- spTransform(malaria_utm, CRS("+proj=longlat +datum=WGS84"))
# add long lat coordinates to malaria dataframe
malaria_village[, c("long", "lat")] <- coordinates(malaria_geo)
malaria_village = st_as_sf(malaria_village, coords = c("long", "lat"),
                           crs=4326)

# specify the bounding box
latLongBox = bbox(malaria_geo)
location = c(latLongBox[1, 1] - 0.05,
             latLongBox[2, 1] - 0.05,
             latLongBox[1, 2] + 0.05,
             latLongBox[2, 2] + 0.05)

# create map with location dots marked on it in
library(rnaturalearth)
GambiaMap <- ne_countries(type = "countries",
                    country = "Gambia",
                    scale = "large", returnclass = "sf")

ggplot() + geom_sf(data=GambiaMap) + 
  geom_sf(data = malaria_village,
          aes(col = prev),
          size = 2) + scale_color_viridis() + theme_void()

```


###  Nimble {-} 

```{r Ex 10.14 nimble model}

Example10_14_Nimble <- nimbleCode({
  
  # Define priors
  sigma ~ T(dt(mu = 0, sigma = 1, df = 1), 0, Inf)
  sigma_sq <- sigma ^ 2
  phi_inv ~ dgamma(shape =  5, rate = 5)
  phi <- 1 / phi_inv

  Sigma[1:N, 1:N] <-
     sigma_sq*(1 + (sqrt(3)*obs_dist_mat[1:N, 1:N])/phi) * exp(-sqrt(3)*obs_dist_mat[1:N, 1:N] / phi) 

    for(i in 1:N){
    mean_S[i] <- b0
    }
  
  S[1:N] ~ dmnorm(mean_S[1:N], cov = Sigma[1:N,1:N])

    for (j in 1:n) {  # by child
    logit(p[j])  <- inprod(b[1:k], X[j,1:k]) + S[index_village[j]]
    y[j] ~ dbern(p[j])
    }

  for (l in 1:k){
    b[l] ~ dnorm(0, sd = 5)
  }
  
  b0 ~ dnorm(0, sd = 5)
  


})

```

```{r Ex 10.14 prepare data nimble model, warning = FALSE, error = FALSE, message = FALSE, results='hide', cache = TRUE}
# distance specification
coords_sf <- sf::st_geometry(malaria_village)
obs_dist_mat <- sf::st_distance(coords_sf)
obs_dist_mat <- units::set_units(obs_dist_mat, km)
obs_dist_mat <- units::drop_units(obs_dist_mat)

# define indicator variables for each village

gambia_df <- mutate(
  gambia,
  id_child = 1:nrow(gambia), # add an id for each child
  value = 1, # this will be needed later for the villages
  id_village = as.numeric(interaction( # add an id for the villages
    x, y, drop = TRUE, lex.order = TRUE
  ))
)  |> 
  tidyr::spread(id_area, value, fill = 0) |>
  rename(area1 = '1', area2 = '2', area3 = '3', area4 = '4', area5 = '5')

## Model specification
# Variables matrix
X <- data.frame(age = scale(gambia_df[,"age"], center = TRUE, scale = FALSE),
                netuse = gambia_df[,"netuse"],
                treated = gambia_df[,"treated"],
                green = scale(gambia_df[,"green"], center = TRUE, scale = FALSE),
                phc = gambia_df[,"phc"],
                area2 = gambia_df[,"area2"],
                area3 = gambia_df[,"area3"],
                area4 = gambia_df[,"area4"],
                area5 = gambia_df[,"area5"]
                )

index_village <- gambia_df[,"id_village"]

n <- nrow(X) # child number
N <- nrow(malaria_village) # number of villages
zeroes <- rep(0, N) # auxiliary vector of zeroes for model
ones <- rep(1, N)

const_list <- list(n = n, # number of childs
                   N = N, # number of villages
                   zeroes = zeroes, # vector of zeroes 
                   prior_max_dist = max(obs_dist_mat)/6, # max dist for phi prior
                   k =  ncol(X),# number of predictors
                   index_village = index_village,
                   Id10 = 10*diag(N)) 

dat_list <- list(y = gambia_df$pos, # malaria positive test 
                 obs_dist_mat = obs_dist_mat, # distance matrix in km
                 X = X # predictors matrix
                 )

init_list <- list(sigma = 0.5, 
                  p = rep(expit(rnorm(1, 0, 1)), n), 
                  phi_inv = 6/max(obs_dist_mat),
                  b = rep(0, ncol(X)), 
                  b0 = rnorm(1, 0, 1),
                  S = rnorm(N, 0, 1))

#init_list <- list(p = runif(n, 0, 1), b = rnorm(ncol(X), 0,  1), b0 = rnorm(1, 0, 1))

Rmodel <-
  nimbleModel(
    Example10_14_Nimble,
    constants = const_list,
    data = dat_list,
    inits = init_list
  )
Rmodel$initializeInfo()
Cmodel <- compileNimble(Rmodel, showCompilerOutput = FALSE)
conf <-
  configureMCMC(Rmodel, monitors = c( "b0", "b", "p", "S", "sigma", "phi"))

# conf$removeSamplers(c('S'))
# conf$addSampler(target = c('S'), type = 'AF_slice')

Rmcmc <- buildMCMC(conf)
Cmcmc <-
  compileNimble(
    Rmcmc,
    project = Cmodel,
    resetFunctions = TRUE,
    showCompilerOutput = TRUE
  )
niters <- 80000
nburnins <- 0.5 * niters
nchains <- 2
nthins <- 14
post_samples <- runMCMC(
  Cmcmc,
  niter = niters,
  nburnin = nburnins,
  thin = nthins,
  nchains = nchains,
  samplesAsCodaMCMC = TRUE,
  summary = TRUE
)
```



```{r Ex 10.14 nimble traceplots ESS }

plot(post_samples$samples[, c("b0")], bty = "n", main = "b0")
plot(post_samples$samples[, c("b[1]")], bty = "n", main = "b[1]")
plot(post_samples$samples[, c("b[2]")], bty = "n", main = "b[2]")
plot(post_samples$samples[, c("b[3]")], bty = "n", main = "b[3]")
plot(post_samples$samples[, c("b[4]")], bty = "n", main = "b[4]")
plot(post_samples$samples[, c("b[5]")], bty = "n", main = "b[5]")
plot(post_samples$samples[, c("b[6]")], bty = "n", main = "b[6]")
plot(post_samples$samples[, c("sigma")], bty = "n", main = "sigma")
plot(post_samples$samples[, c("phi")], bty = "n", main = "phi")
plot(post_samples$samples[, c("S[1]")], bty = "n", main = "S[1]")
plot(post_samples$samples[, c("S[24]")], bty = "n", main = "S[24]")
plot(post_samples$samples[, c("S[54]")], bty = "n", main = "S[54]")
plot(post_samples$samples[, c("p[1]")], bty = "n", main = "p[1]")
plot(post_samples$samples[, c("p[1805]")], bty = "n", main = "p[1805]")

# Get minimum effective size (ESS) and which variable has the min ESS
min(coda::effectiveSize(post_samples$samples))
mcmc_variable_names <- colnames(post_samples$samples$chain1)

mcmc_variable_names[which(coda::effectiveSize(post_samples$samples) == min(coda::effectiveSize(post_samples$samples)))]

```



```{r Ex 10.14 nimble summary, echo = TRUE}

# Extract samples
variables <- c("b0", "b[1]", "b[2]", "b[3]", "b[4]", "b[5]", "b[6]", "b[7]", "b[8]", "b[9]" ,"sigma", "phi")
summary_nimble <- post_samples$summary$all.chains
summary_nimble[variables,]

```



```{r Ex 10.14 nimble summary p, echo = TRUE}
# Plot posterior summary for the spatial random effect by village
post_summary <- post_samples$summary$all.chains

post_sum_S <-
  as.data.frame(post_summary) |>  tibble::rownames_to_column() |>
  filter(str_detect(rowname, "S")) |>
  dplyr::select(rowname, `95%CI_low`, Mean, `95%CI_upp`)  |>
  mutate(village = gsub(".*?([0-9]+).*", "\\1", rowname))

post_sum_S$village <-
  factor(post_sum_S$village , levels = 1:65)

ggplot(data = post_sum_S, aes(x = village)) +
  geom_pointrange(aes(ymin = `95%CI_low`, ymax = `95%CI_upp`, y = Mean)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_discrete(
    breaks = post_sum_S$village[seq(1, length(post_sum_S$village), by = 5)]) +
  theme_classic() +  ylab("") + xlab("village") +
  ggtitle("Posterior summary spatial random effect by village")

```



### Stan {-}

```{r Ex 10.14 load stan, message=FALSE, echo = FALSE, warning=FALSE, message= FALSE}
# load rstan with options
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


```{r engine='bash', comment='', echo = FALSE}
cat functions/Example10_14.stan
```


```{r Ex 10.14 stan data, warning = FALSE, cache = TRUE}

ex.data <- list( n = nrow(gambia), # number of children
                 k = ncol(X), # number of covariates
                 N = N, # number of villages
                 y = gambia$pos, # positive tests
                 dist_matrix = obs_dist_mat, # distance matrix in km
                 X = X, # altitude per village
                 index_village = index_village
                 )

Example10_14Stan  <- stan(
  file = "functions/Example10_14.stan",
  data = ex.data,
  warmup = 15000,
  iter = 30000,
  chains = 2,
  thin = 10,
  pars = c("beta0", "betas","sigma", "phi", "S"),
  include = TRUE
)

```

```{r Ex 10.14 stan traceplots}
#computing WAIC using the package loo

rstan::traceplot(Example10_14Stan, pars = c("beta0","betas","sigma",  "phi"))

```


```{r Ex 10.14 stan summary, echo = TRUE}

# Extract samples
summary_stan <-
  summary(
    Example10_14Stan,
    pars = c("beta0","betas", "sigma", "phi"),
    probs = c(0.025, 0.975)
  )

summary_stan$summary

```



```{r Ex 10.14 stan posterior plots}
S_summary <-
  summary(Example10_14Stan, pars = c("S"))$summary

S_summary_df <- data.frame(S_summary) |>
  tibble::rownames_to_column() |>
  filter(rowname %in% paste0("S[", 1:65, "]")) |>
  mutate(village = 1:65) |>
  dplyr::select(mean,  X2.5., X97.5., village)

S_summary_df$village <-
  factor(S_summary_df$village , levels = 1:65)


  ggplot(S_summary_df, aes(x = village, group = 1)) +
  geom_pointrange(aes(ymin = X2.5., ymax = X97.5., y = mean)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_discrete(
    breaks = S_summary_df$village[seq(1, length(S_summary_df$village), by = 5)]) +
  theme_classic() +  ylab("") + xlab("village") +
  ggtitle("Posterior summary spatial random effect by village")



```

